---
title: "Data Science Midterm"
author: "Nick Williams"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)

theme_set(theme_bw()) # setting a constant theme
```

### Importing data

```{r importing data, message = FALSE}
accel_data <- read_csv("./data/p8105_mtp_data.csv") %>% 
  janitor::clean_names() # importing and cleaning variable names
```

### Cleaning data

```{r normalize or tidy data and creating an hour variable}
accel_data <- accel_data %>% 
  gather(key = "minute", value = "activity_count", activity_1:activity_1440) %>% # converting to long form
  separate(minute, into = c("axe", "minute"), sep = "_") %>% 
  select(-axe) %>% 
  mutate(day = as_factor(day), 
         day = fct_relevel(day, c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                  "Thursday", "Friday", "Saturday")), # re-leveling the day variable
         minute = as.integer(minute), 
         hour = cut(minute, breaks = c(-Inf, 61, 121, 181, 241, 301, 361, 
                                       421, 481, 541, 601, 661, 721, 781, 
                                       841, 901, 961, 1021, 1081, 1141, 1201, 
                                       1261, 1321, 1381, 1441), 
                    labels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 
                               13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24) # creating an hour variable
         ), 
         week_fct = as.factor(week), 
         # creating a collapsed week factor
         week_fct = fct_collapse(week_fct, 
                                 "1 - 5" = c("1", "2", "3", "4", "5"),
                                 "6 - 10" = c("6", "7", "8", "9", "10"), 
                                 "11 - 15" = c("11", "12", "13", "14", "15"),
                                 "16 - 20" = c("16", "17", "18", "19", "20"), 
                                 "21 - 25" = c("21", "22", "23", "24", "25"), 
                                 "26 - 30" = c("26", "27", "28", "29", "30"), 
                                 "31 - 35" = c("31", "32", "33", "34", "35"), 
                                 "36 - 40" = c("36", "37", "38", "39", "40", "41"), 
                                 "41 - 47" = c("42", "43", "44", "45", "46", "47")) 
  )
```

The first steps in this analysis required importing and tidying the data. I transformed the dataset from wide form to long form (using the variables activity_1 to activity_1440) by creating a new variable called minute that records what minute of the day each row corresponds to. I then created a new variable called hour that recorded what hour of the day each observation belonged to. Lastly, I created  another week variable that collapsed the weeks into 5/6 week intervals. The day variable was ordered so that the factor levels followed the actual order of the week.

The cleaned accelerometer dataset contains `r nrow(accel_data)` observations and `r ncol(accel_data)` variables. Variables include the week, day, hour, and minute. Ultimately, the data represents `r nrow(accel_data) / 1440` days or `r as.integer(nrow(accel_data) / 24)` hours worth of data. The data is highly right skewed due to the presence of outliers; as such, in addition to total activity, the median and/or geometric mean are used throughout this analysis. The median and geometric average activity count throughout the entire observation period were `r median(accel_data$activity_count)` and `r round(exp(mean(log(accel_data$activity_count))), 2)` respectively. The modal activity count was 1 with `r accel_data %>% group_by(activity_count) %>% summarize(n = n()) %>% filter(n == max(n)) %>% .$n` observed minutes.

### Analyses

```{r density plots of activity over the weeks, out.width = "65%", fig.asp = .8, fig.align = "center", dpi = 300, warning = FALSE}
density_full <- accel_data %>% 
  mutate(log_actvty_count = log(activity_count)) %>% # creating log activity variable
  ggplot(aes(x = log_actvty_count, color = week_fct)) +
  geom_density(size = 0.75, fill = "grey", alpha = 0.1) + # indicating to make density plot
  viridis::scale_color_viridis(discrete = TRUE, 
                               name = "Week", 
                               option = "cividis") + 
  labs(x = "Log Activity Count", 
       y = "Density") + 
  theme(legend.position = c(.93, .55), # putting legend inside plot
        legend.key.size = unit(.1, "in"),
        legend.title = element_text(size = 10), 
        legend.background = element_blank(), 
        legend.box.background = element_rect(color = "black"))

density_zoom_1 <- accel_data %>% 
  mutate(log_actvty_count = log(activity_count)) %>% 
  ggplot(aes(x = log_actvty_count, color = week_fct)) +
  geom_density(size = 0.75, fill = "grey", alpha = 0.1) + 
  # coord_cartesian is used to zoom in on the x axis
  coord_cartesian(xlim = c(0, 1)) + 
  viridis::scale_color_viridis(discrete = TRUE, 
                               guide = FALSE, 
                               option = "cividis") + 
  labs(caption = "Log activity zoomed between 0 and 1") + 
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank())

density_zoom_2 <- accel_data %>% 
  mutate(log_actvty_count = log(activity_count)) %>% 
  ggplot(aes(x = log_actvty_count, color = week_fct)) +
  geom_density(size = 0.75, fill = "grey", alpha = 0.1) + 
  # once again zooming in on the x axis 
  coord_cartesian(xlim = c(2, 8), ylim = c(0, 0.3)) + 
  viridis::scale_color_viridis(discrete = TRUE, 
                               guide = FALSE, 
                               option = "cividis") +  
  labs(caption = "Log activity zoomed between 2 and 8") + 
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank())

density_full / (density_zoom_1 + density_zoom_2) + 
  plot_annotation(title = "Density plots of log activity count with zoomed axis", 
                  caption = "Data provided by Jeff Goldsmith, PhD")
```

In the above figure, I plotted the distribution of the log activity counts stratified by weekly intervals. The density plot is characterized by two distinct peaks centered around zero and six. Interestingly, the peak centered around zero (indicating minimal activity) begins to decrease as time moves forward while the peak centered around six begins to increase. This is the first indication that the individual's activity may have increased over time. 

```{r activity by day line plot, fig.width = 10, fig.asp = .4, out.width = "100%", dpi = 300, fig.align = "center"}
geo_agg_activity <- accel_data %>% 
  group_by(day) %>% 
  mutate(log_actvty = log(activity_count)) %>% 
  summarize(geo_avg_actvty = exp(mean(log_actvty))) %>% # calculating geometric mean
  ggplot(aes(x = day, y = geo_avg_actvty)) + 
  geom_point(size = 4, color = "#A35E60") + 
  geom_line(group = 1) + 
  labs(title = "Geometric average activity by the day of the week", 
       x = "Day of the week", 
       y = "Geometric average activity count")

total_agg_activity <- accel_data %>% 
  group_by(day) %>% 
  summarize(total_actvty = sum(activity_count)) %>%
  ggplot(aes(x = day, y = total_actvty)) + 
  geom_point(size = 4, color = "#A35E60") + 
  geom_line(group = 1) + 
  labs(title = "Total activity by the day of the week", 
       x = "Day of the week", 
       y = "Total activity count")

total_agg_activity + geo_agg_activity + 
  plot_annotation(caption = "Data provided by Jeff Goldsmith, PhD")
```

I also explored the effect of weekday on activity. The figure above shows total activity and the geometric mean of activity across the days of the week aggregated across all weeks. The lowest total activity occurs on Tuesdays while the highest occurs on Friday's. Interestingly, however, based on the geometric mean, we'd expect that the lowest activity during any week would occur on Sunday instead of Tuesday. 
  
```{r median activity by day and hour, fig.asp = .7, out.width = "70%", fig.align = "center", dpi = 300}
med_actvty_all_days <- accel_data %>% #created a separate dataset with median activity aggregated
  group_by(hour) %>% 
  summarize(med_activity = median(activity_count)) %>% 
  mutate(day = as.factor("All days"), 
         hour = as.double(hour))

accel_data %>% 
  mutate(hour = as.double(hour)) %>% 
  group_by(hour, day) %>% 
  summarize(med_activity = median(activity_count)) %>% 
  ggplot(aes(x = hour, y = med_activity, color = day)) +
  geom_line(size = 1, alpha = 0.5) + 
  # plotting aggregated data as a dashed line
  geom_line(data = med_actvty_all_days, lty = "dashed", size = 1.5) + 
  viridis::scale_color_viridis(name = "Day of the week", 
                               discrete = TRUE, 
                               option = "viridis",
                               # adding a legend break for the aggregate level
                               breaks = c("All days", "Sunday", "Monday", "Tuesday",
                                          "Wednesday", "Thursday", "Friday", "Saturday")) + 
  scale_x_continuous(limits = c(0, 24), 
                     breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)) + 
  scale_y_continuous(minor_breaks = seq(0, 300, 25)) + 
  labs(title = "Activity count throughout the day stratified by day of the week",
       x = "Hour of the Day (24 hour clock)", 
       y = "Median Activity Count", 
       caption = "Data provided by Jeff Goldsmith, PhD") + 
  theme(legend.position = c(.05, 1), 
        legend.justification = c(0, 1.1), 
        legend.key.size = unit(.1, "in"),
        # putting legend inside plot
        legend.title = element_text(size = 10), 
        legend.background = element_blank(), 
        legend.box.background = element_rect(color = "black"))
```

Furthermore, I investigated the effect of time of day on activity. The above figure shows trends in median hourly activity stratified by weekday as well as across all days. Activity appears to be relatively constant during daytime hours. However, there are peaks in activity on Thursday and Friday between 6 p.m. and 9 p.m. 

```{r total activity across weeks, out.width = "100%", fig.width = 10, fig.asp = 0.65, fig.align = "center", dpi = 300, message = FALSE}
total_actvty <- accel_data %>% 
  group_by(week) %>% 
  summarize(total_actvty = sum(activity_count)) %>% 
  ggplot(aes(x = week, y = total_actvty)) + 
  geom_line() + # making line plot
  geom_smooth(color = "#A35E60", se = FALSE) + # adding a smoothing line for trend
  labs(y = "Total Activity") + 
  theme(axis.title.x = element_blank())

total_actvty_week_day <- accel_data %>% 
  group_by(week, day) %>% 
  summarize(total_actvty = sum(activity_count)) %>% #
  ggplot(aes(x = week, y = total_actvty)) + 
  facet_grid(~ day) + # creating a facet across days of the week
  geom_line() + 
  labs(x = "Week", 
       y = "Total Activity") + 
  geom_smooth(color = "#A35E60", se = FALSE) + 
  theme(strip.background = element_blank(), 
        strip.text = element_text(color = "black")
  )

total_actvty / total_actvty_week_day  + 
  plot_annotation(title = "Trends in total activity over the observation period", 
                  caption = "Data provided by Jeff Goldsmith, PhD")
```

Lastly, I analyzed how total activity changed across the whole observation period. The above figure shows the trend in activity across all 47 weeks alone and stratified by day of the week. As time increases, it appears that total activity does increase as well. It also appears that total activity had more variation towards the beginning of the study and became more consistent towards the end. An increase in activity is observed across all days over the observation period. The trend of increase follows a similar pattern across all the days. However, for Saturday after week 35 there is a decrease in activity. Furthermore, it appears the rate of increase tapers off towards the end of the observation period. 


